---
title: "Introduction to MaxMin"
author: "Martin R. Smith"
date: today
format: html
vignette: >
  %\VignetteIndexEntry{Introduction to MaxMin}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library("MaxMin")
```

The **Max-Min Diversity Problem** (MMDP) asks: given a fixed set of *N* candidate
items and a distance between every pair, choose *m* items so that the
closest pair in the selection is as far apart as possible.
The objective — often written T~k~ — is the minimum pairwise distance within
the chosen subset; a larger T~k~ means a more spread-out, representative
selection.

This problem arises naturally wherever you want a diverse sample from a fixed
pool: selecting field-survey sites to cover a landscape, picking biological
specimens for sequencing that span the available genetic diversity, or choosing
a representative subset of protein structures from a database.

## Installation

```{r}
#| label: install
#| eval: false
install.packages("MaxMin")
```

For the exact solver, also install **highs**:

```{r}
#| label: install-highs
#| eval: false
install.packages("highs")
```

## Quick start

`eurodist` is a built-in R `dist` object containing road distances (km)
between 21 European cities.

```{r}
#| label: quickstart
data(eurodist)

# Select 4 maximally dispersed cities. The default ensemble uses random
# starts, so set a seed for a reproducible selection.
set.seed(1)
idx <- Gonzalez(eurodist, n = 4L)
MinDist(eurodist, idx)

as.matrix(eurodist)[idx, idx]
```

`Gonzalez()` returned the indices of the four cities whose nearest-neighbour
distance within the selection is largest.
`MinDist()` reports that value explicitly.

## Methods at a glance

MaxMin provides four solvers and two utilities:

| Function | Quality | Speed | Stochastic? |
|---|---|---|---|
| `Gonzalez()` | Good (2-approximation) | Very fast | No |
| `DropAddTS()` | High (≈ 99 % optimal) | Fast | No |
| `GraspPR()` | Highest | Moderate | Yes (`seed =`) |
| `ExactMaxMin()` | Optimal (NP-hard) | Slow | No |
| `MinDist()` | Scoring only | Instant | No |

The rest of this vignette walks through each in turn.

---

## Gonzalez: fast greedy selection

The **Gonzalez** (1985) algorithm builds the selection greedily:
start from a seed point, then repeatedly add whichever unselected point is
farthest from the current selection.
This greedy rule guarantees a 2-approximation to the optimal T~k~ and runs
in O(*N* · *m*) time.

By default `Gonzalez()` runs an ensemble of three `"random_furthest"` starts
and returns whichever pass produced the highest T~k~ (a best-of-three
selection). The random starts pivot on points drawn with the session RNG, so set
a seed (`set.seed()`) for a reproducible selection. Pass a `pivots` vector for
more (or your own) random starts — its length sets the number of starts. The
deterministic anchors (`"centroid"`, `"peripheral"`, and the costlier O(*N*²)
anchors) are opt-in: pass a character vector to `seed` to choose a custom
ensemble, or a single name for one strategy.

```{r}
#| label: gonzalez-ensemble
set.seed(1)
idx_ens <- Gonzalez(eurodist, n = 6L)   # default: best of three random starts

# Which seeding strategy won?
attr(idx_ens, "winning_strategy")
```

You can also request a specific seeding strategy or a custom ensemble subset:

```{r}
#| label: gonzalez-seeds
idx_diam <- Gonzalez(eurodist, n = 6L, seed = "diameter")
MinDist(eurodist, idx_diam)

idx_anti <- Gonzalez(eurodist, n = 6L, seed = "anti_medoid")
MinDist(eurodist, idx_anti)
```

When only pairwise distances between objects are known — there are no
coordinates to work from — and computing or storing all N×N distances at once
would be too expensive (e.g. large phylogenetic trees, where each tree-to-tree
distance requires its own calculation), pass `Gonzalez()` a *function* in place
of the distance matrix. The function takes one index `i` and returns the
distances from object `i` to all N objects — one column of the matrix. Gonzalez
calls it only *n* times, so the full N×N matrix is never built; you must supply
`N` (the number of objects), since it cannot be read off a function.

```{r}
#| label: gonzalez-oracle
dm <- as.matrix(eurodist)
col_of <- function(i) dm[, i]       # one column on demand
Gonzalez(col_of, n = 6L, N = nrow(dm), seed = 1L)
```

(Here the column simply indexes a matrix we already have, to keep the example
self-contained; the point is that `col_of` could instead *compute* each column
from trees, sequences, or any metric with no coordinate embedding.) Only an
integer `seed` (a fixed start index) or the default peripheral seed is
available on this path — the richer ensemble anchors need the whole matrix.

---

## DropAdd tabu search

The **DropAdd** heuristic (Porumbel, Hao & Glover 2011) refines an initial
selection by alternating *drop* and *add* moves guided by a FIFO tabu list
that prevents short cycles.
It is fully deterministic and reproducible, and typically reaches ≈ 99 % of
the optimal T~k~.

```{r}
#| label: dropadd
res_da <- DropAddTS(eurodist, m = 6L, max_no_improve = 500L)

labels(eurodist)[res_da$indices]
res_da$objective     # T_k achieved
res_da$iters         # iterations completed
res_da$time_s        # wall-clock seconds
```

`max_no_improve` is the main stopping knob: the algorithm terminates after
this many consecutive iterations without an improvement in T~k~.
For coordinate data where *N* is large enough to make the distance matrix
infeasible (roughly N > 46 000), use `DropAddTSPoints()`, which accepts a
coordinate matrix directly.

---

## GRASP with path relinking

**GRASP + path relinking** (Resende et al. 2010) combines a *randomised*
greedy construction phase with extended local search and then refines an elite
set of good solutions by interpolating between elite-pair trajectories
(*path relinking*).
It achieves the highest T~k~ of the three heuristics, at a proportionally
higher cost.
Because the construction phase is stochastic, always set `seed` for
reproducibility:

```{r}
#| label: grasp
res_gr <- GraspPR(eurodist, m = 6L, max_no_improve = 50L, seed = 42L)

labels(eurodist)[res_gr$indices]
res_gr$objective
res_gr$pr_calls   # path-relinking calls performed
```

`max_no_improve` controls how many consecutive non-improving GRASP iterations
trigger termination; `time_budget_s` is available as an absolute time cap for
interactive or batch use.

---

## Comparing methods on a simulated example

To see how the methods relate visually, we generate 50 points in two
dimensions and select *m* = 8 from each.

```{r}
#| label: sim-data
set.seed(42)
pts <- matrix(rnorm(100), ncol = 2)   # 50 points, 2 dimensions
d50 <- dist(pts)
m   <- 8L
```

```{r}
#| label: compare-run
idx_gonz <- Gonzalez(d50, n = m)
res_da50 <- DropAddTS(d50, m = m, max_no_improve = 500L)
res_gr50 <- GraspPR(d50, m = m, max_no_improve = 50L, seed = 42L)
```

Even a small difference in T~k~ can correspond to a meaningfully more
dispersed selection.

```{r}
#| label: compare-scores
scores <- c(
  Gonzalez  = MinDist(d50, idx_gonz),
  DropAddTS = res_da50$objective,
  GraspPR   = res_gr50$objective
)
round(scores, 3)
```

Plotting the selections against the point cloud makes the differences
concrete.
When two methods select the same index, their symbols overlap; the T~k~ table
above captures the quality distinction even when the visual overlap is high.

```{r}
#| label: compare-plot
#| fig-width: 6.5
#| fig-height: 6
#| fig-cap: "Selections returned by each method on 50 random 2-D points
#|   (m = 8). Coloured symbols mark selected points; grey circles are the
#|   full candidate set. A small jitter separates symbols at shared indices."

methods <- list(
  Gonzalez  = idx_gonz,
  DropAddTS = res_da50$indices,
  GraspPR   = res_gr50$indices
)
cols <- c(Gonzalez = "#E41A1C", DropAddTS = "#377EB8", GraspPR = "#4DAF4A")
pchs <- c(Gonzalez = 24L, DropAddTS = 21L, GraspPR = 22L)
# Small per-method shift so coincident selections remain distinguishable
dx   <- c(Gonzalez = 0, DropAddTS =  0.04, GraspPR = -0.04)
dy   <- c(Gonzalez = 0, DropAddTS = -0.04, GraspPR =  0.04)

plot(pts, pch = 1L, col = "grey75", asp = 1L,
     xlab = "x", ylab = "y",
     main = "MaxMin method comparison")

for (nm in names(methods)) {
  sel <- methods[[nm]]
  points(
    pts[sel, 1L] + dx[nm],
    pts[sel, 2L] + dy[nm],
    pch = pchs[nm], col = cols[nm], bg = cols[nm], cex = 1.6
  )
}

legend("topright",
       legend = paste0(names(methods), "  (Tₖ = ", round(scores, 3), ")"),
       pch = pchs, col = cols, pt.bg = cols, pt.cex = 1.4, bty = "n")
```

---

## Exact solution

For small instances (roughly N ≤ 25–30), `ExactMaxMin()` solves the problem
to proven optimality via a node-packing integer programme (Sayyady & Fathi
2016), using the **highs** solver.

```{r}
#| label: exact-data
set.seed(1L)
pts30 <- matrix(rnorm(60L), ncol = 2L)
d30   <- dist(pts30)
```

```{r}
#| label: exact
#| eval: !expr requireNamespace("highs", quietly = TRUE)
res_ex <- ExactMaxMin(d30, m = 6L, time_budget_s = 30L)

res_ex$proven      # TRUE  ⟹  objective is the global optimum
res_ex$objective

# Compare to the greedy heuristic on the same instance
res_gonz30 <- Gonzalez(d30, n = 6L)
c(exact    = res_ex$objective,
  Gonzalez = MinDist(d30, res_gonz30))
```

`$proven = TRUE` certifies that no selection can achieve a higher T~k~.
`ExactMaxMin()` is NP-hard; instances with more than ~30 candidates require a
time budget or should be tackled with the heuristics above.

---

## Scoring

### `MinDist()`

`MinDist()` computes the T~k~ objective for any index set.
It accepts a `dist` object, a square distance matrix, or a coordinate matrix
via the `points` argument:

```{r}
#| label: MinDist
MinDist(d50, idx_gonz)                               # from dist
MinDist(as.matrix(d50), idx_gonz)                    # from square matrix
MinDist(points = pts, idx = idx_gonz)                # from coordinates
```

---

## When to use which method

| Scenario | Recommended |
|---|---|
| Speed matters most, N up to a few thousand | `Gonzalez()` (ensemble default) |
| Deterministic, reproducible refinement | `DropAddTS()` |
| Best quality, set `seed` for reproducibility | `GraspPR(seed = ...)` |
| N > 46 000 (distance matrix infeasible) | `DropAddTSPoints()` or `Gonzalez(points = ...)` |
| Arbitrary metric with no coordinate embedding | `Gonzalez(<column function>, N = ...)` |
| Proven optimum, N ≤ ~ 25–30, **highs** installed | `ExactMaxMin()` |

The heuristics are complementary:
`Gonzalez()` is O(*N* · *m*) and deterministic — an instant first result.
`DropAddTS()` is deterministic and reproducible (no RNG), typically reaching
≈ 99 % of optimal; the `seed` parameter governs only the RNG used for
tie-breaking, not the algorithm itself.
`GraspPR()` is stochastic and usually edges out `DropAddTS()` on T~k~, but
requires a `seed` for reproducibility.

## Related packages

The CRAN package [**maximin**](https://CRAN.R-project.org/package=maximin)
constructs continuous space-filling designs by generating new points.
