---
title: "Introduction to MaxMin"
author: "Martin R. Smith"
date: today
format: html
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa.csl
vignette: >
  %\VignetteIndexEntry{Introduction to MaxMin}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library("MaxMin")
```

The MaxMin package selects a subset that represents a fixed pool of *N* items,
based on one of two complementary objectives:

The **Max-Min Diversity Problem** (MMDP, the discrete *p*-dispersion objective)
selects $k$ elements such that the minimum distance between any pair of selected elements is as large as possible; the chosen elements are maximally separated.
This can reward selections that leave the interior of the set unrepresented.

This objective is suited to defining a representative sample from a fixed
pool: picking biological specimens for sequencing that span available diversity,
or choosing a representative subset of protein structures from a database.

The **discrete *k*-centre problem** selects $k$ elements such that the maximum
distance from any element in the original set to a selected element is as small
as possible.
In ensuring that each point has a nearby representative, this objective can 
select points that reflect a central compromise, rather than selections that
are closer to more local points.

This objective is useful when selecting centres that represent each point in
a dataset: for example, siting fire stations to guarantee that all buildings
can be reached within a given response time.

An approximate selection that satisfies both objectives within a factor of two
of their respective optima can be attained by a greedy farthest-first algorithm
[@Gonzalez1985], though the exact optima typically differ; dispersion spreads to
the extremes, whereas covering reaches into the interior.
MaxMin provides approximate and exact solvers for each objective.


## Installation

```{r}
#| label: install
#| eval: false
install.packages("MaxMin")
```

## Quick start

Our examples employ a built-in R `dist` object that contains road distances
(in km) between 21 European cities.

```{r}
#| label: quickstart
# Load the `eurodist` dist object
data("eurodist")

# Set a seed for a reproducible selection
set.seed(1)

# Select 4 maximally dispersed cities
ffPick <- FarFirst(4L, eurodist)

# View distances between selected cities
as.matrix(eurodist)[ffPick, ffPick]

# Quickly extract the minimum distance for a given selection
MinDist(eurodist, ffPick)
```

`FarFirst()` returned the indices of the four cities whose nearest-neighbour
distance within the selection is largest.
`MinDist()` reports that value explicitly.

## Methods at a glance

MaxMin provides solvers for each objective:

| Function | Objective | Quality | Speed | Stochastic? |
|---|---|---|---|---|
| `FarFirst()` | both | Good (2-approximation) | Very fast | No |
| `DropAdd()` | MMDP | High (≈ 99 % optimal) | Fast | No |
| `Grasp()` | MMDP | Highest | Moderate | Yes (`set.seed()`) |
| `ExactMaxMin()` | MMDP | Optimal (NP-hard) | Slow | No |
| `KCentre()` | k-centre | Near-optimal (CDSh) | Fast | No |
| `ExactKCentre()` | k-centre | Optimal (NP-hard) | Slow | No |


To compute the score for an arbitrary selection of points under each objective,
use `MinDist()` (MMDP) and `KCentreRadius()` (k-centre).


## Fast greedy selection

The @Gonzalez1985 algorithm builds the selection greedily:
start from a seed point, then repeatedly add whichever unselected point is
farthest from the current selection.
This greedy rule guarantees a 2-approximation to the optimal T~k~ and runs
in O(*N* · *m*) time.

The choice of seed point influences the quality of the selection. Peripheral
seeds are more likely than central seeds to represent extremes of the data,
and hence to feature in the optimal selection.  `FarFirst()` supports several
methods for identifying starting points for the greedy search.

The default is a two-step strategy that starts from a random selects a point at random, then starts greedy search from
the point furthest from 

By default `FarFirst()` runs eight `"random_furthest"` starts, each of which
takes a randomly selected point, moves to the point furthest from it, and begins
the farthest-first sweep from there. The results of the pass with the highest
T~k~ are returned.

```{r}
#| label: gonzalez-random
set.seed(1)
FarFirst(6L, eurodist)   # default: best of eight random starts
```

The number of random starts can be configured via the `nSeeds` argument.

```{r}
#| label: gonzalez-ensemble
set.seed(1)
# Fewer starts provides a faster run, but the solution found may be inferior
FarFirst(6L, eurodist, nSeeds = 2L)
```


Other strategies to select peripheral seeds are available via the `strategy`
argument; these are described at `PickPoint()`.
The best solution from the requested strategies is used.

```{r}
#| label: gonzalez-seeds
ffPick <- FarFirst(6L, eurodist, strategy = c("diameter", "anti_medoid"))
MinDist(eurodist, ffPick)

attr(ffPick, "winning_strategy")
```

With large datasets of points that are not associated with natural coordinates,
it may not be feasible to compute and store all N×N distances.
In such cases, a distance function may take the place of the distance matrix.
This function is passed one index `i` at a time, and must return the distance
from `i` to each other point.
`N`, the number of objects, must also be specified.

```{r}
#| label: gonzalez-oracle
data("USArrests")
arrestTypes <- USArrests[, c("Murder", "Assault", "Rape")]
StateDist <- function(i) {
  diffs <- sweep(arrestTypes, 2, unlist(arrestTypes[i, ]), "-")
  sqrt(rowSums(diffs ^ 2))
}
idx <- FarFirst(4L, StateDist, N = nrow(arrestTypes), strategy = 1L)
arrestTypes[idx, ]
```


## DropAdd tabu search

The **DropAdd** heuristic [@Porumbel2011] refines an initial
selection by alternately dropping and adding points from the selection; it 
typically reaches ≈ 99 % of the optimal T~k~.

```{r}
#| label: dropadd
daPick <- DropAdd(6L, eurodist, plateau = 500L)

daPick
labels(eurodist)[daPick]
```

The algorithm terminates after `plateau` iterations do not improve T~k~.
Where *N* is too large for a distance matrix to fit in memory 
(roughly N > 46 000), pass a coordinate matrix via `DropAdd(points = ...)`.


## GRASP with path relinking

**GRASP + path relinking** [@Resende2010] combines a randomised greedy
construction phase with extended local search, and then refines an "elite"
set of good solutions by interpolating between elite-pair trajectories
(path relinking).
It achieves the highest T~k~ of the three heuristics, at a proportionally
higher cost.

```{r}
#| label: grasp
set.seed(0)
grPick <- Grasp(6L, eurodist, plateau = 50L)
grPick
labels(eurodist)[grPick]
attr(grPick, "pr_calls")   # path-relinking calls performed
```

`plateau` controls how many consecutive non-improving GRASP iterations
trigger termination; `maxSeconds` is available as an absolute time cap.


## Comparing methods on a simulated example

To see how the methods relate visually, we generate 50 points in two
dimensions and select *k* = 8 from each.

```{r}
#| label: sim-data
set.seed(1) # Seed selected such that FarFirst < DropAdd < Grasp
pts <- matrix(rnorm(100), ncol = 2)   # 50 points, 2 dimensions
d50 <- dist(pts)
k   <- 8L
```

```{r}
#| label: compare-run
set.seed(1) # Seed selected such that FarFirst < DropAdd < Grasp
ffPick <- FarFirst(k, d50)
da50Pick <- DropAdd(k, d50, plateau = 500L)
gr50Pick <- Grasp(k, d50, plateau = 50L)
```

Even a small difference in T~k~ can correspond to a meaningfully more
dispersed selection.

```{r}
#| label: compare-scores
scores <- c(
  FarFirst  = attr(ffPick, "score"),
  DropAdd = attr(da50Pick, "score"),
  Grasp   = attr(gr50Pick, "score")
)
round(scores, 3)
```

Plotting the selections against the point cloud makes the differences
concrete.
When two methods select the same index, their symbols overlap; the T~k~ table
above captures the quality distinction even when the visual overlap is high.

```{r}
#| label: compare-plot
#| fig-width: 6.5
#| fig-height: 6
#| fig-cap: "Selections returned by each method on 50 random 2-D points
#|   (k = 8). Coloured symbols mark selected points; grey circles are the
#|   full candidate set."

methods <- list(
  FarFirst  = ffPick,
  DropAdd = da50Pick,
  Grasp   = gr50Pick
)
cols <- c(FarFirst = "#E41A1C", DropAdd = "#377EB8", Grasp = "#4DAF4A")
pchs <- c(FarFirst = 1L, DropAdd = 3L, Grasp = 4L)

plot(pts, pch = 1L, col = "grey75", asp = 1L,
     xlab = "x", ylab = "y", frame.plot = FALSE,
     main = "MaxMin method comparison")

for (nm in names(methods)) {
  sel <- methods[[nm]]
  points(pts[sel, 1L], pts[sel, 2L], pch = pchs[nm], col = cols[nm],
         cex = 1.6)
}

legend("topleft",
       legend = paste0(names(methods), "  (T\U2096 = ", round(scores, 3), ")"),
       pch = pchs, col = cols, pt.bg = cols, pt.cex = 1.4, bty = "n")
```


## Exact solution

For small instances (roughly N ≤ 25–30), `ExactMaxMin()` solves the problem
to proven optimality via a node-packing integer programme [@Sayyady2016],
using the **highs** solver (which we must first install).

```{r}
#| label: install-highs
#| eval: false
install.packages("highs")
```

```{r}
#| label: exact-data
set.seed(1L)
pts30 <- matrix(rnorm(60L), ncol = 2L)
d30   <- dist(pts30)
```

```{r}
#| label: exact
#| eval: !expr requireNamespace("highs", quietly = TRUE)
exPick <- ExactMaxMin(6L, d30, maxSeconds = 30L)

attr(exPick, "proven")      # TRUE  ⟹  objective is the global optimum
attr(exPick, "score")

# Compare to the greedy heuristic on the same instance
ff30Pick <- FarFirst(6L, d30)
c(exact    = attr(exPick, "score"),
  farFirst = MinDist(d30, ff30Pick))
```

`$proven = TRUE` certifies that no selection can achieve a higher T~k~.
`ExactMaxMin()` is NP-hard; it is wise to set a time budget once the number
of points exceeds ~30 candidates, or to switch to a heuristic method.


## Scoring

`MinDist()` computes the T~k~ objective for any index set.
It accepts a `dist` object, a square distance matrix, or a coordinate matrix
via the `points` argument:

```{r}
#| label: MinDist
MinDist(d50, ffPick)                               # from dist
MinDist(as.matrix(d50), ffPick)                    # from square matrix
MinDist(points = pts, idx = ffPick)                # from coordinates
```


## Covering: the k-centre problem

The above methods seek to spread the selection such that its members are
mutually far apart. The k-centre problem instead minimizes the covering radius
$R$, the largest distance from any point to its nearest chosen centre, such that
no point of the pool is left far from a representative
[@Gonzalez1985; @Hochbaum1985], typically resulting in selections that reach
further into the interior of a sample.


### Heuristic approach

`FarFirst()` is the quickest approximation to the K-centres problem.
The CDSh algorithm implemented in `KCentres()` gives a more sophisticated
heuristic [@GarciaDiaz2019; @GarciaDiaz2017]; its solutions are typically within 
1–3.5 % of the optimum, compared to ~10% for `FarFirst()`.

```{r}
#| label: kcentre
centres <- KCentre(4L, eurodist)
labels(eurodist)[centres]
centres
```

`KCentreRadius()` scores any centre set by its covering radius (lower is
better). CDSh covers at least as tightly as the Gonzalez 2-approximation
baseline:

```{r}
#| label: kcentre-radius
ff <- FarFirst(4L, eurodist, strategy = "peripheral")
c(KCentre  = KCentreRadius(eurodist, centres),
  FarFirst = KCentreRadius(eurodist, ff))
```

Like `MinDist()`, `KCentreRadius()` accepts a `points` coordinate matrix,
allowing it to score a selection on a set of points whose distance matrix would
be too large to fit in memory.

### Exact solver

For small instances, `ExactKCentre()` finds the optimal solution, using a 
minimum-set-cover integer program approach akin to `ExactMaxMin()`, and using
again the **highs** solver.

```{r}
#| label: exact-kcentre
#| eval: !expr requireNamespace("highs", quietly = TRUE)
kc <- ExactKCentre(4L, eurodist)
kc
attr(kc, "proven")      # TRUE  ⟹  radius is the global covering optimum
```

The covering optimum is sometimes attained by fewer centres (once every
point is covered, extra centres cannot lower the radius); `indices` then has
length below the requested number, and the reported `radius` is still the proven optimum.
`ExactKCentre()` is NP-hard, so — like `ExactMaxMin()` — it is a ground-truth
reference for small instances, not a scalable method.

The dispersion and covering optima differ even on this small example: dispersion
selects cities at the rim of the map, while covering pulls inward to keep every
city near a centre.

```{r}
#| label: dispersion-vs-covering
#| eval: !expr requireNamespace("highs", quietly = TRUE)
disp <- ExactMaxMin(4L, eurodist)$indices
labels(eurodist)[disp]  # dispersion: pushed to the extremes
labels(eurodist)[kc]    # covering: pulled toward the interior
```


## When to use which method

For **dispersion** (spread the selection; maximize T~k~):

| Scenario | Recommended |
|---|---|
| Speed matters most | `FarFirst()` (ensemble default) |
| Deterministic, good quality valued | `DropAdd()` |
| Best quality, `set.seed()` for reproducibility | `Grasp()` |
| N > 46 000 (distance matrix infeasible) | `DropAdd(points = ...)` or `FarFirst(points = ...)` |
| Arbitrary metric with no coordinate embedding | `FarFirst(<column function>, N = ...)` |
| Proven optimum, N ≤ ~ 25–30, **highs** installed | `ExactMaxMin()` |
| Score a selection's T~k~ | `MinDist()` |

For **covering** (minimise the radius; no point far from a centre):

| Scenario | Recommended |
|---|---|
| Near-optimal covering, fast and deterministic | `KCentre()` (CDSh) |
| Quick 2-approximation baseline | `FarFirst(strategy = "peripheral")` |
| Proven optimum, small N, **highs** installed | `ExactKCentre()` |
| Score a centre set's covering radius (matrix-free at large N) | `KCentreRadius(points = ...)` |


## Related problems

`MaxMin` selects a subset from a given set of elements.
Several established packages solve neighbouring objectives:

- **k-medoids / k-median** selects elements that minimize the mean distance from
  each element to its nearest centre.
  Implementations include:
  * [`cluster::pam()`](https://cran.r-project.org/package=cluster): generates
  the full *O(N²)* dissimilarity matrix, and hence caps at *n* ≤ 65 536;
  * [`banditpam`](https://cran.r-project.org/package=banditpam), a matrix-free
  $O(N \log N)$ implementation restricted to coordinate data;
  * `cluster::clara()`  (PAM / FastPAM / FasterPAM);
  * [`ClusterR::Cluster_Medoids()`](https://cran.r-project.org/package=ClusterR).

- **k-means** ([`stats::kmeans()`](https://rdrr.io/r/stats/kmeans.html)) selects
  elements so as to minimize the within-cluster sum of squares around centres
  that are coordinate means, not data points; as such, it applies only to
  Euclidean coordinates. k-means++ 
  ([`TreeDist::KMeansPP()`](
   https://ms609.github.io/TreeDist/reference/KMeansPP.html)) initializes its
   selection using D²-weighted seeding, a randomized relative of `FarFirst()`'s
   farthest-first traversal.

- [`maximin`](https://cran.r-project.org/package=maximin) solves the related
  design problem of adding *new* points at positions that maximize the minimum
  inter-point distance.

## References
