---
title: "Introduction to MaxMin"
author: "Martin R. Smith"
date: today
format: html
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa.csl
vignette: >
  %\VignetteIndexEntry{Introduction to MaxMin}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library("MaxMin")
```

The **Max-Min Diversity Problem** (MMDP) asks: given a fixed set of *N* candidate
items and a distance between every pair, choose *m* items so that the
closest pair in the selection is as far apart as possible.
The objective — often written T~k~ — is the minimum pairwise distance within
the chosen subset; a larger T~k~ means a more spread-out, representative
selection.

This problem arises naturally wherever you want a diverse sample from a fixed
pool: selecting field-survey sites to cover a landscape, picking biological
specimens for sequencing that span the available genetic diversity, or choosing
a representative subset of protein structures from a database.

## Installation

```{r}
#| label: install
#| eval: false
install.packages("MaxMin")
```

## Quick start

`eurodist` is a built-in R `dist` object containing road distances (km)
between 21 European cities.

```{r}
#| label: quickstart
data(eurodist)

# Set a seed for a reproducible selection
set.seed(1)

# Select 4 maximally dispersed cities
idx <- Gonzalez(eurodist, n = 4L)
MinDist(eurodist, idx)

# View distances between chosen cities
as.matrix(eurodist)[idx, idx]
```

`Gonzalez()` returned the indices of the four cities whose nearest-neighbour
distance within the selection is largest.
`MinDist()` reports that value explicitly.

## Methods at a glance

MaxMin provides four solvers, and a minimum-distance calculator:

| Function | Quality | Speed | Stochastic? |
|---|---|---|---|
| `Gonzalez()` | Good (2-approximation) | Very fast | No |
| `DropAdd()` | High (≈ 99 % optimal) | Fast | No |
| `GraspPR()` | Highest | Moderate | Yes (`seed =`) |
| `ExactMaxMin()` | Optimal (NP-hard) | Slow | No |
| `MinDist()` | Scoring only | Instant | No |

We now walk through each of these in turn.


## Gonzalez: fast greedy selection

The Gonzalez [-@Gonzalez1985] algorithm builds the selection greedily:
start from a seed point, then repeatedly add whichever unselected point is
farthest from the current selection.
This greedy rule guarantees a 2-approximation to the optimal T~k~ and runs
in O(*N* · *m*) time.

By default `Gonzalez()` runs three `"random_furthest"` starts, each beginning
from the point furthest from a randomly selected 'pivot' point, and returns
whichever pass produced the highest T~k~.

```{r}
#| label: gonzalez-random
set.seed(1)
picks <- Gonzalez(eurodist, n = 6L)   # default: best of three random starts
```

More random starts, or pivots of your own devising, can be accomplished by a
vector of pivot points to a `pivots` vector.

```{r}
#| label: gonzalez-ensemble
set.seed(1)
nCities <- attr(eurodist, "Size")
# Use 5 random starts
pivots <- sample(nCities, 5)
picks <- Gonzalez(eurodist, n = 6L, pivots = pivots)
```


Peripheral seeds may also be selected by deterministic methods: one or more
such methods can be selected via the `seed` argument. The best solution found
will be returned.

```{r}
#| label: gonzalez-seeds
picks <- Gonzalez(eurodist, n = 6L, seed = c("diameter", "anti_medoid"))
MinDist(eurodist, picks)

attr(picks, "winning_strategy")
```

When only pairwise distances between objects are known — there are no
coordinates to work from — and computing or storing all N×N distances at once
would be too expensive, a function may be passed in place
of the distance matrix. This function will be passed one index `i`, and should
returns the distances from object `i` to all N objects — one column of the
matrix.
`Gonzalez()` calls it $n$ times, so the full N×N matrix is never built.
`N`, the number of objects, must be supplied.

```{r}
#| label: gonzalez-oracle
data("USArrests")
arrestTypes <- USArrests[, c("Murder", "Assault", "Rape")]
StateDist <- function(i) {
  diffs <- sweep(arrestTypes, 2, unlist(arrestTypes[i, ]), "-")
  sqrt(rowSums(diffs ^ 2))
}
idx <- Gonzalez(StateDist, n = 4L, N = nrow(arrestTypes), seed = 1L)
arrestTypes[idx, ]
```


## DropAdd tabu search

The **DropAdd** heuristic [@Porumbel2011] refines an initial
selection by alternately dropping and adding points from the selection; it 
typically reaches ≈ 99 % of the optimal T~k~.

```{r}
#| label: dropadd
picksDA <- DropAdd(eurodist, m = 6L, plateau = 500L)

labels(eurodist)[picksDA]
attr(picksDA, "objective")  # T_k achieved
attr(picksDA, "iters")      # iterations completed
attr(picksDA, "time_s")     # wall-clock seconds
```

The algorithm terminates after `plateau` iterations do not improve T~k~.
Where *N* is too large for a distance matrix to fit in memory 
(roughly N > 46 000), `DropAddPoints()` accepts a coordinate matrix directly.


## GRASP with path relinking

**GRASP + path relinking** [@Resende2010] combines a *randomised*
greedy construction phase with extended local search and then refines an elite
set of good solutions by interpolating between elite-pair trajectories
(*path relinking*).
It achieves the highest T~k~ of the three heuristics, at a proportionally
higher cost.
Because the construction phase is stochastic, always set `seed` for
reproducibility:

```{r}
#| label: grasp
res_gr <- GraspPR(eurodist, m = 6L, plateau = 50L, seed = 42L)

labels(eurodist)[res_gr$indices]
res_gr$objective
res_gr$pr_calls   # path-relinking calls performed
```

`plateau` controls how many consecutive non-improving GRASP iterations
trigger termination; `timeBudgetS` is available as an absolute time cap for
interactive or batch use.


## Comparing methods on a simulated example

To see how the methods relate visually, we generate 50 points in two
dimensions and select *m* = 8 from each.

```{r}
#| label: sim-data
set.seed(42)
pts <- matrix(rnorm(100), ncol = 2)   # 50 points, 2 dimensions
d50 <- dist(pts)
m   <- 8L
```

```{r}
#| label: compare-run
idx_gonz <- Gonzalez(d50, n = m)
res_da50 <- DropAdd(d50, m = m, plateau = 500L)
res_gr50 <- GraspPR(d50, m = m, plateau = 50L, seed = 42L)
```

Even a small difference in T~k~ can correspond to a meaningfully more
dispersed selection.

```{r}
#| label: compare-scores
scores <- c(
  Gonzalez  = MinDist(d50, idx_gonz),
  DropAdd = attr(res_da50, "objective"),
  GraspPR   = res_gr50$objective
)
round(scores, 3)
```

Plotting the selections against the point cloud makes the differences
concrete.
When two methods select the same index, their symbols overlap; the T~k~ table
above captures the quality distinction even when the visual overlap is high.

```{r}
#| label: compare-plot
#| fig-width: 6.5
#| fig-height: 6
#| fig-cap: "Selections returned by each method on 50 random 2-D points
#|   (m = 8). Coloured symbols mark selected points; grey circles are the
#|   full candidate set. A small jitter separates symbols at shared indices."

methods <- list(
  Gonzalez  = idx_gonz,
  DropAdd = res_da50,
  GraspPR   = res_gr50$indices
)
cols <- c(Gonzalez = "#E41A1C", DropAdd = "#377EB8", GraspPR = "#4DAF4A")
pchs <- c(Gonzalez = 24L, DropAdd = 21L, GraspPR = 22L)
# Small per-method shift so coincident selections remain distinguishable
dx   <- c(Gonzalez = 0, DropAdd =  0.04, GraspPR = -0.04)
dy   <- c(Gonzalez = 0, DropAdd = -0.04, GraspPR =  0.04)

plot(pts, pch = 1L, col = "grey75", asp = 1L,
     xlab = "x", ylab = "y",
     main = "MaxMin method comparison")

for (nm in names(methods)) {
  sel <- methods[[nm]]
  points(
    pts[sel, 1L] + dx[nm],
    pts[sel, 2L] + dy[nm],
    pch = pchs[nm], col = cols[nm], bg = cols[nm], cex = 1.6
  )
}

legend("topright",
       legend = paste0(names(methods), "  (Tₖ = ", round(scores, 3), ")"),
       pch = pchs, col = cols, pt.bg = cols, pt.cex = 1.4, bty = "n")
```


## Exact solution

For small instances (roughly N ≤ 25–30), `ExactMaxMin()` solves the problem
to proven optimality via a node-packing integer programme [@Sayyady2016],
using the **highs** solver.

First we must install **highs**:

```{r}
#| label: install-highs
#| eval: false
install.packages("highs")
```

```{r}
#| label: exact-data
set.seed(1L)
pts30 <- matrix(rnorm(60L), ncol = 2L)
d30   <- dist(pts30)
```

```{r}
#| label: exact
#| eval: !expr requireNamespace("highs", quietly = TRUE)
res_ex <- ExactMaxMin(d30, m = 6L, timeBudgetS = 30L)

res_ex$proven      # TRUE  ⟹  objective is the global optimum
res_ex$objective

# Compare to the greedy heuristic on the same instance
res_gonz30 <- Gonzalez(d30, n = 6L)
c(exact    = res_ex$objective,
  Gonzalez = MinDist(d30, res_gonz30))
```

`$proven = TRUE` certifies that no selection can achieve a higher T~k~.
`ExactMaxMin()` is NP-hard; instances with more than ~30 candidates require a
time budget or should be tackled with the heuristics above.


## Scoring

### `MinDist()`

`MinDist()` computes the T~k~ objective for any index set.
It accepts a `dist` object, a square distance matrix, or a coordinate matrix
via the `points` argument:

```{r}
#| label: MinDist
MinDist(d50, idx_gonz)                               # from dist
MinDist(as.matrix(d50), idx_gonz)                    # from square matrix
MinDist(points = pts, idx = idx_gonz)                # from coordinates
```


## When to use which method

| Scenario | Recommended |
|---|---|
| Speed matters most, N up to a few thousand | `Gonzalez()` (ensemble default) |
| Deterministic, reproducible refinement | `DropAdd()` |
| Best quality, set `seed` for reproducibility | `GraspPR(seed = ...)` |
| N > 46 000 (distance matrix infeasible) | `DropAddPoints()` or `Gonzalez(points = ...)` |
| Arbitrary metric with no coordinate embedding | `Gonzalez(<column function>, N = ...)` |
| Proven optimum, N ≤ ~ 25–30, **highs** installed | `ExactMaxMin()` |

The heuristics are complementary:
`Gonzalez()` is O(*N* · *m*) and deterministic — an instant first result.
`DropAdd()` is deterministic and reproducible (no RNG), typically reaching
≈ 99 % of optimal; the `seed` parameter governs only the RNG used for
tie-breaking, not the algorithm itself.
`GraspPR()` is stochastic and usually edges out `DropAdd()` on T~k~, but
requires a `seed` for reproducibility.

## Related packages

The CRAN package [**maximin**](https://CRAN.R-project.org/package=maximin)
constructs continuous space-filling designs by generating new points.

## References
