---
title: "Introduction to MaxMin"
author: "Martin R. Smith"
date: today
format: html
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa.csl
vignette: >
  %\VignetteIndexEntry{Introduction to MaxMin}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library("MaxMin")
```

The MaxMin package selects a subset that represents a fixed pool of *N* items,
based on one of two complementary objectives:

The **Max-Min Diversity Problem** (MMDP, the discrete *p*-dispersion objective)
selects $k$ elements such that the minimum distance between any pair of selected elements is as large as possible; the chosen elements are maximally separated.
This can reward selections that leave the interior of the set unrepresented.

This objective is suited to defining a representative sample from a fixed
pool: picking biological specimens for sequencing that span available diversity,
or choosing a representative subset of protein structures from a database.

The **discrete *k*-centre problem** selects $k$ elements such that the maximum
distance from any element in the original set to a selected element is as small
as possible.
In ensuring that each point has a nearby representative, this objective can 
select points that reflect a central compromise, rather than selections that
are closer to more local points.

This objective is useful when selecting centres that represent each point in
a dataset: for example, siting fire stations to guarantee that all buildings
can be reached within a given response time.

An approximate selection that satisfies both objectives within a factor of two
of their respective optima can be attained by a greedy farthest-first algorithm
[@Gonzalez1985], though the exact optima typically differ; dispersion spreads to
the extremes, whereas covering reaches into the interior.
MaxMin provides approximate and exact solvers for each objective.


## Installation

```{r}
#| label: install
#| eval: false
install.packages("MaxMin")
```

## Quick start

Our examples employ a built-in R `dist` object that contains road distances
(in km) between 21 European cities.

```{r}
#| label: quickstart
# Load the `eurodist` dist object
data("eurodist")

# Set a seed for a reproducible selection
set.seed(1)

# Select 4 maximally dispersed cities
ffPick <- FarFirst(4L, eurodist)

# View distances between selected cities
as.matrix(eurodist)[ffPick, ffPick]

# Quickly extract the minimum distance for a given selection
MinDist(eurodist, ffPick)
```

`FarFirst()` returned the indices of the four cities whose nearest-neighbour
distance within the selection is largest.
`MinDist()` reports that value explicitly.

## Methods at a glance

MaxMin provides solvers for each objective:

| Function | Objective | Quality | Speed | Stochastic? |
|---|---|---|---|---|
| `FarFirst()` | both | Good (2-approximation) | Very fast | No |
| `DropAdd()` | MMDP | High (≈ 99 % optimal) | Fast | No |
| `Grasp()` | MMDP | Highest | Moderate | Yes (`set.seed()`) |
| `ExactMaxMin()` | MMDP | Optimal (NP-hard) | Slow | No |
| `KCentre()` | k-centre | Near-optimal (CDSh) | Fast | No |
| `ExactKCentre()` | k-centre | Optimal (NP-hard) | Slow | No |


To compute the score for an arbitrary selection of points under each objective,
use `MinDist()` (MMDP) and `KCentreRadius()` (k-centre).


## Fast greedy selection

The @Gonzalez1985 algorithm builds the selection greedily:
start from a seed point, then repeatedly add whichever unselected point is
farthest from the current selection.
This greedy rule guarantees a 2-approximation to the optimal T~k~ and runs
in O(*N* · *m*) time.

By default `FarFirst()` runs three `"random_furthest"` starts, each beginning
from the point furthest from a randomly selected 'pivot' point, and returns
whichever pass produced the highest T~k~.

```{r}
#| label: gonzalez-random
set.seed(1)
ffPick <- FarFirst(6L, eurodist)   # default: best of eight random starts
```

More random starts can be requested via the `nseeds` argument.

```{r}
#| label: gonzalez-ensemble
set.seed(1)
ffPick <- FarFirst(6L, eurodist, nseeds = 12L)
```


Peripheral seeds may also be selected by deterministic strategies: one or more
such strategies can be selected via the `strategy` argument. The best solution found
will be returned.

```{r}
#| label: gonzalez-seeds
ffPick <- FarFirst(6L, eurodist, strategy = c("diameter", "anti_medoid"))
MinDist(eurodist, ffPick)

attr(ffPick, "winning_strategy")
```

When only pairwise distances between objects are known — there are no
coordinates to work from — and computing or storing all N×N distances at once
would be too expensive, a function may be passed in place
of the distance matrix. This function will be passed one index `i`, and should
returns the distances from object `i` to all N objects — one column of the
matrix.
`FarFirst()` calls it $k$ times, so the full N×N matrix is never built.
`N`, the number of objects, must be supplied.

```{r}
#| label: gonzalez-oracle
data("USArrests")
arrestTypes <- USArrests[, c("Murder", "Assault", "Rape")]
StateDist <- function(i) {
  diffs <- sweep(arrestTypes, 2, unlist(arrestTypes[i, ]), "-")
  sqrt(rowSums(diffs ^ 2))
}
idx <- FarFirst(4L, StateDist, N = nrow(arrestTypes), strategy = 1L)
arrestTypes[idx, ]
```


## DropAdd tabu search

The **DropAdd** heuristic [@Porumbel2011] refines an initial
selection by alternately dropping and adding points from the selection; it 
typically reaches ≈ 99 % of the optimal T~k~.

```{r}
#| label: dropadd
daPick <- DropAdd(eurodist, k = 6L, plateau = 500L)

daPick
labels(eurodist)[daPick]
```

The algorithm terminates after `plateau` iterations do not improve T~k~.
Where *N* is too large for a distance matrix to fit in memory 
(roughly N > 46 000), pass a coordinate matrix via `DropAdd(points = ...)`.


## GRASP with path relinking

**GRASP + path relinking** [@Resende2010] combines a *randomised*
greedy construction phase with extended local search and then refines an elite
set of good solutions by interpolating between elite-pair trajectories
(*path relinking*).
It achieves the highest T~k~ of the three heuristics, at a proportionally
higher cost.
Because the construction phase is stochastic, call `set.seed()` before
`Grasp()` for a reproducible run:

```{r}
#| label: grasp
set.seed(42)
grPick <- Grasp(eurodist, k = 6L, plateau = 50L)
grPick
labels(eurodist)[grPick]
attr(grPick, "pr_calls")   # path-relinking calls performed
```

`plateau` controls how many consecutive non-improving GRASP iterations
trigger termination; `maxSeconds` is available as an absolute time cap for
interactive or batch use.


## Comparing methods on a simulated example

To see how the methods relate visually, we generate 50 points in two
dimensions and select *k* = 8 from each.

```{r}
#| label: sim-data
set.seed(42)
pts <- matrix(rnorm(100), ncol = 2)   # 50 points, 2 dimensions
d50 <- dist(pts)
k   <- 8L
```

```{r}
#| label: compare-run
ffPick <- FarFirst(k, d50)
da50Pick <- DropAdd(d50, k = k, plateau = 500L)
set.seed(42)
gr50Pick <- Grasp(d50, k = k, plateau = 50L)
```

Even a small difference in T~k~ can correspond to a meaningfully more
dispersed selection.

```{r}
#| label: compare-scores
scores <- c(
  FarFirst  = MinDist(d50, ffPick),
  DropAdd = attr(da50Pick, "score"),
  Grasp   = attr(gr50Pick, "score")
)
round(scores, 3)
```

Plotting the selections against the point cloud makes the differences
concrete.
When two methods select the same index, their symbols overlap; the T~k~ table
above captures the quality distinction even when the visual overlap is high.

```{r}
#| label: compare-plot
#| fig-width: 6.5
#| fig-height: 6
#| fig-cap: "Selections returned by each method on 50 random 2-D points
#|   (k = 8). Coloured symbols mark selected points; grey circles are the
#|   full candidate set. A small jitter separates symbols at shared indices."

methods <- list(
  FarFirst  = ffPick,
  DropAdd = da50Pick,
  Grasp   = gr50Pick
)
cols <- c(FarFirst = "#E41A1C", DropAdd = "#377EB8", Grasp = "#4DAF4A")
pchs <- c(FarFirst = 24L, DropAdd = 21L, Grasp = 22L)
# Small per-method shift so coincident selections remain distinguishable
dx   <- c(FarFirst = 0, DropAdd =  0.04, Grasp = -0.04)
dy   <- c(FarFirst = 0, DropAdd = -0.04, Grasp =  0.04)

plot(pts, pch = 1L, col = "grey75", asp = 1L,
     xlab = "x", ylab = "y",
     main = "MaxMin method comparison")

for (nm in names(methods)) {
  sel <- methods[[nm]]
  points(
    pts[sel, 1L] + dx[nm],
    pts[sel, 2L] + dy[nm],
    pch = pchs[nm], col = cols[nm], bg = cols[nm], cex = 1.6
  )
}

legend("topright",
       legend = paste0(names(methods), "  (Tₖ = ", round(scores, 3), ")"),
       pch = pchs, col = cols, pt.bg = cols, pt.cex = 1.4, bty = "n")
```


## Exact solution

For small instances (roughly N ≤ 25–30), `ExactMaxMin()` solves the problem
to proven optimality via a node-packing integer programme [@Sayyady2016],
using the **highs** solver.

First we must install **highs**:

```{r}
#| label: install-highs
#| eval: false
install.packages("highs")
```

```{r}
#| label: exact-data
set.seed(1L)
pts30 <- matrix(rnorm(60L), ncol = 2L)
d30   <- dist(pts30)
```

```{r}
#| label: exact
#| eval: !expr requireNamespace("highs", quietly = TRUE)
exPick <- ExactMaxMin(d30, k = 6L, maxSeconds = 30L)

exPick$proven      # TRUE  ⟹  objective is the global optimum
exPick$objective

# Compare to the greedy heuristic on the same instance
ff30Pick <- FarFirst(6L, d30)
c(exact    = exPick$objective,
  farFirst = MinDist(d30, ff30Pick))
```

`$proven = TRUE` certifies that no selection can achieve a higher T~k~.
`ExactMaxMin()` is NP-hard; instances with more than ~30 candidates require a
time budget or should be tackled with the heuristics above.


## Scoring

### `MinDist()`

`MinDist()` computes the T~k~ objective for any index set.
It accepts a `dist` object, a square distance matrix, or a coordinate matrix
via the `points` argument:

```{r}
#| label: MinDist
MinDist(d50, ffPick)                               # from dist
MinDist(as.matrix(d50), ffPick)                    # from square matrix
MinDist(points = pts, idx = ffPick)                # from coordinates
```


## Covering: the k-centre problem

Every method above pursues *dispersion* — it spreads the selection so its
members are mutually far apart. The **k-centre** problem instead pursues
*coverage*: it minimises the covering radius *R*, the largest distance from any
point to its nearest chosen centre, so that no point of the pool is left far
from a representative [@Gonzalez1985; @Hochbaum1985]. Where dispersion reaches
for the extremes, covering reaches into the interior, so on the same data the
two optima are generally different selections.

### `KCentre()`: near-optimal covering

`KCentre()` chooses centres with the deterministic CDSh heuristic
[@GarciaDiaz2019; @GarciaDiaz2017], which typically lands within 1–3.5 % of the
optimum — an order of magnitude tighter than the Gonzalez 2-approximation that
`FarFirst()` gives for this objective.

```{r}
#| label: kcentre
centres <- KCentre(eurodist, k = 4L)
labels(eurodist)[centres]
centres
```

`KCentreRadius()` scores any centre set by its covering radius (lower is
better). CDSh covers at least as tightly as the Gonzalez 2-approximation
baseline:

```{r}
#| label: kcentre-radius
ff <- FarFirst(4L, eurodist, strategy = "peripheral")
c(KCentre  = KCentreRadius(eurodist, centres),
  FarFirst = KCentreRadius(eurodist, ff))
```

Like `MinDist()`, `KCentreRadius()` also accepts a `points` coordinate matrix,
in which case it never materialises the full *N × N* matrix — so it can score a
selection well past the size at which the solvers' distance matrix would fit in
memory.

### `ExactKCentre()`: proven covering optimum

For small instances, `ExactKCentre()` solves the covering problem to proven
optimality with a sequence of minimum-set-cover integer programs — the covering
dual of `ExactMaxMin()`'s node-packing programme — warm-started from the
`KCentre()` radius and bisected down to the smallest feasible radius. Like
`ExactMaxMin()`, it uses the **highs** solver.

```{r}
#| label: exact-kcentre
#| eval: !expr requireNamespace("highs", quietly = TRUE)
kc <- ExactKCentre(eurodist, k = 4L, progress = FALSE)
kc
kc$proven      # TRUE  ⟹  radius is the global covering optimum
```

The covering optimum is sometimes attained by fewer centres (once every
point is covered, extra centres cannot lower the radius); `indices` then has
length below the requested number, and the reported `radius` is still the proven optimum.
`ExactKCentre()` is NP-hard, so — like `ExactMaxMin()` — it is a ground-truth
reference for small instances, not a scalable method.

The dispersion and covering optima differ even on this small example: dispersion
selects cities at the rim of the map, while covering pulls inward to keep every
city near a centre.

```{r}
#| label: dispersion-vs-covering
#| eval: !expr requireNamespace("highs", quietly = TRUE)
disp <- sort(ExactMaxMin(eurodist, k = 4L)$indices)
labels(eurodist)[disp]              # dispersion: pushed to the extremes
labels(eurodist)[sort(kc$indices)]  # covering: pulled toward the interior
```


## When to use which method

For **dispersion** (spread the selection; maximise T~k~):

| Scenario | Recommended |
|---|---|
| Speed matters most, N up to a few thousand | `FarFirst()` (ensemble default) |
| Deterministic, reproducible refinement | `DropAdd()` |
| Best quality, `set.seed()` for reproducibility | `Grasp()` |
| N > 46 000 (distance matrix infeasible) | `DropAdd(points = ...)` or `FarFirst(points = ...)` |
| Arbitrary metric with no coordinate embedding | `FarFirst(<column function>, N = ...)` |
| Proven optimum, N ≤ ~ 25–30, **highs** installed | `ExactMaxMin()` |
| Score a selection's T~k~ | `MinDist()` |

For **covering** (minimise the radius; no point far from a centre):

| Scenario | Recommended |
|---|---|
| Near-optimal covering, fast and deterministic | `KCentre()` (CDSh) |
| A quick 2-approximation baseline | `FarFirst(strategy = "peripheral")` |
| Proven optimum, small N, **highs** installed | `ExactKCentre()` |
| Score a centre set's covering radius (matrix-free at large N) | `KCentreRadius(points = ...)` |

The dispersion heuristics are complementary:
`FarFirst()` is O(*N* · *k*) and deterministic — an instant first result.
`DropAdd()` is deterministic and reproducible (no RNG): ties are broken by
smallest index, so a given instance always yields the same selection,
typically reaching ≈ 99 % of optimal.
`Grasp()` is stochastic and usually edges out `DropAdd()` on T~k~; call
`set.seed()` before it for a reproducible run.

## Related packages

The CRAN package [**maximin**](https://CRAN.R-project.org/package=maximin)
constructs continuous space-filling designs by generating new points.

## References
